Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-07 text/paper2/ms.tex
moved potential stuff to sims

Another approach to using halo stars as potential measures is to exploit the non-random nature of halo. For example, \cite{law10} used N-body simulations of the disruption of the Sgr dwarf to simultaneously model the debris and underlying Milky Way potential. By varying parameters of the Milky Way potential, they found ``best-fit'' parameters by comparing the properties of observed Sgr stars to their simulated debris over a grid of potential parameters. The computational costs of running N-body simulations limited their search to a grid of values and forced the authors to fix many other parameters (e.g., properties of the disk and bulge). Nevertheless, they were able to constrain the shape of the Milky Way potential out to $\sim$70 kpc and found that the best-fitting halo has a prolate, triaxial shape elongated in the Galactic Z direction, orthogonal to the plane of the disk. Though an unlikely orientation for the halo --- e.g., \cite{debattista13} find that the disk of the Milky Way would not remain stable in such a configuration --- LM10 showed that the data is at a state where such inference can be attempted. The computational costs associated with N-body modeling has motivated many methods that approximately model tidal streams.

@@

The potential is taken to be a three-component sum of a Miyamoto-Nagai disk \citep{}, Hernquist bulge \citep[spheroid;][]{}, and a triaxial, logarithmic halo \citep[e.g.,][]{law10}:
\begin{align}
	&\Phi_{\rm disk}(R,z) = -\frac{GM_{\rm disk}}{\sqrt{R^2 + (a + \sqrt{z^2 + b^2})^2}}\\
	&\Phi_{\rm spher}(r) = -\frac{GM_{\rm spher}}{r + c}\\
	&\Phi_{\rm halo}(x,y,z) = v_{h}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + r_h^2)
\end{align}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis
ratios ($q_1$, $q_2$) and orientation of the halo with respect to the
baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}
The total potential then is just
\begin{equation}
	\Phi_{\rm tot} = \Phi_{\rm disk} + \Phi_{\rm spher} + \Phi_{\rm halo}\label{eq:lm10}.
\end{equation}

@@

\section{Experiments}
In what follows, we apply this model to ``observations'' of the N-body simulations described in Section~\ref{sec:sims}. For all tests in this article, we model the potential of the parent galaxy with the same functional form used in the simulations (Equation~\ref{}).

Initially we assume the present-day mass of the progenitor, $M_p$, and velocity dispersion, $\sigma_v$, have been measured perfectly and neglect mass-loss from the progenitor.

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{/Users/adrian/projects/streams/plots/paper2/potentials.pdf}
\caption{ TODO }\label{fig:potential}
\end{center}
\end{figure}

When recovering the potential, we hold fixed the disk and spheroid parameters (see Table~\ref{tbl:potential}), along with two halo parameters: the scale radius, $r_h$, and one of the flattening components in the x-y plane, $q_2$. These halo parameters are degenerate with combinations of the other parameters --- e.g., setting $q_1=2.0$, $q_2=1.0$, $\phi=0^\circ$ is equivalent to setting $q_1=1.0$, $q_2=2.0$, $\phi=90^\circ$. The priors on the remaining halo parameters are taken to be uniform over a conservative domain of realistic values: for $v_h$, 100-200~kms$^{-1}$ corresponds to a range in solar circular velocities from $\sim$210-250 kms$^{-1}$ (holding other parameters fixed); the range in axis ratios allow for prolate, oblate, and mixed triaxiality; and $\phi$ is restricted to $\pm45^\circ$ around the value measured by LM10 ($\phi = 97^\circ$). Figure~\ref{fig:potential} shows equipotential slices in the Galactic X-Z plane (Y=0) and Y-Z plane (X=0) for a few choices of  $q_1$,  $q_z$, and $\phi$ while holding all other parameters fixed, as described in the figure caption.

@@

\subsection{\gaia + \spitzer}
4 stars in Sgr-like stream

>>>>

Another approach to using halo stars as potential measures is to exploit the non-random nature of halo. For example, \citet[][LM10]{law10} used N-body simulations of the disruption of the Sgr dwarf to simultaneously model the debris and underlying Milky Way potential. By varying parameters of the Milky Way potential, they found ``best-fit'' parameters by comparing the properties of observed Sgr stars to their simulated debris over a grid of potential parameters. The computational costs of running N-body simulations limited their search to a grid of values and forced the authors to fix many other parameters (e.g., properties of the disk and bulge). Nevertheless, they were able to constrain the shape of the Milky Way potential out to $\sim$70 kpc and found that the best-fitting halo has a prolate, triaxial shape elongated in the Galactic Z direction, orthogonal to the plane of the disk. Though an unlikely orientation for the halo --- e.g., \cite{debattista13} find that the disk of the Milky Way would not remain stable in such a configuration --- LM10 showed that the data is at a state where such inference can be attempted. The computational costs associated with N-body modeling has motivated many methods that approximately model tidal streams.

@@

The potential is taken to be a three-component sum of a Miyamoto-Nagai disk \citep{}, Hernquist bulge \citep[spheroid;][]{}, and a triaxial, logarithmic halo \citep[e.g.,][]{law10}:
\begin{align}
	&\Phi_{\rm disk}(R,z) = -\frac{GM_{\rm disk}}{\sqrt{R^2 + (a + \sqrt{z^2 + b^2})^2}}\\
	&\Phi_{\rm spher}(r) = -\frac{GM_{\rm spher}}{r + c}\\
	&\Phi_{\rm halo}(x,y,z) = v_{h}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + r_h^2)
\end{align}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis
ratios ($q_1$, $q_2$) and orientation of the halo with respect to the
baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}
The total potential then is just
\begin{equation}
	\Phi_{\rm tot} = \Phi_{\rm disk} + \Phi_{\rm spher} + \Phi_{\rm halo}\label{eq:lm10}.
\end{equation}
Figure~\ref{fig:potential} shows equipotential slices in the Galactic X-Z plane (Y=0) and Y-Z plane (X=0) for a few choices of  $q_1$,  $q_z$, and $\phi$ while holding all other parameters fixed, as described in the figure caption.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/potentials.pdf}
\caption{ TODO }\label{fig:potential}
\end{center}
\end{figure}

@@

\section{Experiments}
In what follows, we apply the model described above to ``observations'' of the N-body simulations described in Section~\ref{sec:sims}. For all tests in this article, we model the potential of the parent galaxy with the same functional form used in the simulations (Equation~\ref{eq:lm10}). When recovering the potential, we hold fixed the disk and spheroid parameters (see Table~\ref{tbl:potential}), along with two halo parameters: the scale radius, $r_h$, and one of the flattening components in the x-y plane, $q_2$. These halo parameters are degenerate with combinations of the other parameters --- e.g., setting $q_1=2.0$, $q_2=1.0$, $\phi=0^\circ$ is equivalent to setting $q_1=1.0$, $q_2=2.0$, $\phi=90^\circ$. The priors on the remaining halo parameters are taken to be uniform over a conservative domain of realistic values: for $v_h$, 100-200~kms$^{-1}$ corresponds to a range in solar circular velocities from $\sim$210-250 kms$^{-1}$ (holding other parameters fixed); the range in axis ratios allow for prolate, oblate, and generic triaxiality; and $\phi$ is restricted to $\pm45^\circ$ around the value measured by LM10 ($\phi = 97^\circ$).

@@

\subsection{Perfect data}

\subsection{\gaia + \spitzer}
8 stars in Sgr-like stream

We assume the present-day mass of the progenitor, $M_p$, and velocity dispersion, $\sigma_v$, have been measured perfectly and neglect mass-loss from the progenitor.
